The impact of cosmic variance on constraining the mass and occupation fraction of dark matter halos hosting Lyman-$\alpha$ emitters at $z \sim 3$

### 2014-04-15 paper/ms.tex
correcting typos and small changes to paper

\begin{abstract}
%
  We study the impact of cosmic variance in constraining the mass and
  occupation fraction of dark matter halos hosting \ly Emitting
  Galaxies at high redshift. We use an N-body simulation to construct
  mock fields with the same typical size of observed fields at
  $z=3.1$ to match the observed number density distribution across
  fields and the angular correlation function. In our model a dark
  matter halo with mass in the range $M_{\rm min}<M_{\rm h}<M_{\rm
    max}$ can only host one detectable LAE with a probability $0.1
  \leq f_{\rm occ}\leq 1.0$.  Our analysis shows that the clustering
  and number density information are insufficient to impose a tight
  constraint on the occupation fraction. On the other hand, the
  minimum mass and maximum mass can be constrained to the range
  $M_{\rm  max}<10^{12}$\hMsun\ and $10^{10.2}\hMsun\leq M_{\rm
    min}\leq 10^{11.5}\hMsun$. Additionally, we find that the
  consistent models have a narrow mass range, $\Delta M \equiv \log_{10}M_{\rm max} -
  \log_{10} M_{\rm min}$, smaller than $1.0$ dex, with a majority in
  the range $\Delta M<0.5$ dex. These results fall into constraints
  already presented in the literature. However, the wide range of
  values for the occupation fraction and the narrow mass range $\Delta
  M$ are novel results allowed by takinf into account the influence of
  cosmic variance on the statistics derived from observations.  
\end{abstract}

@@

Capitalizing LAEs observations requires an understanding of
their place in the context of a given structure formation model in a
cosmological context. Under the current paradigm the dominant matter
content of the Universe is to be found in dark matter (DM) and each
galaxy is thought to be hosted by larger dark matter structure known
as a halo. \citep{Peebles1980,SpringelNature05}.

@@

Some theoretical attempts to solve this problem using an  ab-initio
approach. They start from the DM halo population to infer the
intrinsic star formation rates and \ly a luminosities. From these
values they estimate the amount of \ly photons that
escape each galaxy and compute the observed luminosity for each
galaxy. These models can predict different observables including: the
luminosity function, correlation function and the equivalent width
distributions. Such modelling has been implemented in semi-analytic
models \citep{Garel2012,Orsi2012} and  full N-body
hydrodynamical simulations \citep{Laursen2007, Dayal2009,
  ForeroRomero2011, Yajima2012}.

@@

In this paper we investigate the impact of cosmic variance in
constraining the mass and occupation fraction of halos hosting LAEs.
We do this by constructing mock catalogs from cosmological simulations.
Our method populates each halo in the simulation with a LAE without
predicting a \ly  luminosity. This bypasses all the physical
uncertainties associated to star formation and radiative transfer.
We build mock surveys following the geometry of the observed fields by
\cite{Yamada2012} to compared them  against observations in terms of
the number density distribution and the angular correlation
function. This allows us to find a range of parameters of our model
that are consistent with observations while taking into account
cosmic variance.

@@

In the next subsections we describe in detail the four key
elements int our work-flow. First, we present the 
observations we take as a benchmark. Second, we describe the main
characteristics of the N-body simulation and the halo catalogs we
use. Third, we list the important parameters of the simplified
model that we use to populate the halo catalogs with LAEs. Finally, we
describe the statistical tests we adopt to compare observations and mocks.

@@

The narrow band filter used in the survey is
centered at $4977$ \AA~with  $77$ \AA~width, corresponding to the
redshift range $z=3.062$-$3.125$ and $41$ \hMpc comoving scale for the
detection of the Lyman-$\alpha$ line centered at $z=3.09$. The authors
reported a total  $2161$  LAEs with an observed equivalent width, in
the observer frame, larger than $190$ \AA~over a total survey area of
$2.42$ deg$^{2}$ that includes 12 sub-fields,  this corresponds to an
average surface number density of $0.24\pm 0.01$ arcmin$^{-2}$.  The
distribution of the surface number densities is shown in Figure
\ref{fig:number_density}.

@@

The second benchmark is the angular correlation function
(ACF). Unfortunately, \cite{Yamada2012} do  not report an ACF measurement for
their fields. Instead, we use the results by   \cite{Ouchi2008} who
reported the ACF is over a region of $1$ deg$^2$ over the SXDS
field.

@@

We construct a total $5\times 7 \times 6=210$ sub-volumes from a snapshot in the Bolshoi
simulation. In each mock field a LAE is assigned to the position of a
dark matter halo if the halo mass is in the range allowed by the model
$M_{\rm min}<M_{\rm h}<M_{\rm max}$ and a random variable taken from n
homogeneous distribution $0\leq \xi<1$ is smaller than the occupation
fraction $\xi<f_{\rm occ}$.

@@

The lower limit for the parameter $M_{\rm min}$ is set by the minimum
occupation fraction we decide to consider. At $M_{\rm
  min}=10^{10}$\hMsun the halo number density around that mass range
is $\sim 10$ times higher than the observational constraints
for LAEs. This means that models in that mass range and an occupation
fraction $f_{\rm  occ}=0.1$ have the possibility to be compatible with
observations.  Similarly, wxploring occupation fractions on the order
of $f_{\rm occ}=0.01$ only makes sense for halo populations in the
mass range of $10^{9}$\hMsun which are abundant enough to reproduce
the observational constraint on the number density. In our case, this
is a mass range unresolved by the Bolshoi simulation and therefore
is not considered in the present study.

@@

The observed and mock ACF are fit to a power-law function:
\begin{equation}
\omega(\theta) = \left(\frac{\theta}{\theta_{0}}\right)^{-\beta}, 
\label{eq:fitting}
\end{equation}
%
where $\theta_0$ and $\beta$ are free parameters. The fit is done
using a least square minimization procedure. For each mock field we
obtain a covariance matrix that gives us the uncertainty in the
parameters $\beta$ and $\theta_0$.  For a model we derive a mean
value and a dispersion computed from the fit of all mock fields. We
consider that a mocdel is consistent with observations if the two
parameters $\beta$ and $\theta_0$ are equal within a $1$-$\sigma$
range.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.46\linewidth,angle=0]{Fig7_corr_params.pdf}  
\includegraphics[width=0.46\linewidth,angle=0]{Fig7_mmin_vs_theta.pdf} 
\end{center}
\caption{Left panel. Values for the free parameters $\theta_{0}$ and $\beta$
in the fitting formula (Eq. \ref{eq:fitting}) for the angular
correlation function. Blue dots correspond to simulations and the
green line to observations by \citet{Ouchi2008,Ouchi2010}. The error
bars in the theoretical data correspond to the quadratic average of
the fitting uncertainties for each mock survey. Right panel. Values of
$\theta_{0}$  and $\log_{10}(M_{\rm min}/M_{\odot} h^{-1})$. Blue and
green point represents the models where  $\log_{10}(M_{\rm max}/M_{\odot}
h^{-1})<12.0$ and  $\log_{10}(M_{\rm min}/M_{\odot} h^{-1})>12.0$
respectively. .  The horizontal gray band represents the observational
constraints on $theta_{0}$ established by
\citet{Ouchi2008,Ouchi2010}. Green points has been displaced by
$0.01$ dex of its original value to avoid overlapping between blue and
green points.} \label{fig:correlation_parameters}   
\end{figure*}

@@

Figure \ref{fig:halos} allows us to understand why only a specific range of
models ${\mathcal M}$ can be expected to be consistent with
observations. From Figure \ref{fig:halos} we can read that models with
a minimum mass $M_{\rm min}>10^{12}$\hMsun always have a surface
number density lower than the observational constrain, making them
incompatible with observations; there are simply too few halos
compared to observed LAEs. The opposite is true in models with $M_{\rm
  min}<10^{11.0}$\hMsun, which have a surface number density larger
observations. In those cases the maximum mass $M_{\rm max}$ and the
occupation fraction $f_{\rm occ}<1.0$  can be tuned in order to lower
the halo number density to match observations.

@@

The models in the third family are barely consistent with the
constraints from the ACF. These models have another particular
feature: their minimum mass is exactly $M_{\rm min}=10^{10.9}\hMsun$
and $M_{\rm max}>10^{12.0}\hMsun$ . They have values of $\beta$
close to $1.0$ (instead of the observational value of $0.8$) and
$\theta_0$ on the range of $13^{\prime}$ (the mean observational
value is close to $8^{\prime}$) but are considered consistent
because their large error bars barely overlap with the observational
expectation (see figure \ref{fig:correlation_parameters}).
Because these models are are barely compatible with observations and
are a minority of the consistent models, we exclude them from the
discussion.

\begin{figure*}
\begin{center}
\includegraphics[width=0.45\linewidth,angle=0]{Fig6_mass.pdf}
\hspace{5mm}
\includegraphics[width=0.45\linewidth,angle=0]{Fig6_f_occ.pdf}
\end{center}
\caption{$M_{\rm min}$-$M_{\rm max}$ (left) and $M_{\rm
    min}$-$f_{\rm occ}$ (right) planes showing the models fulfilling both
   constraints on the maximal number of consistent mocks and the
  angular correlation function from \citet{Ouchi2008,Ouchi2010}.   
  \label{fig:restriction_mock_and_f_occ_corr}} 
\end{figure*}

@@

There are two interesting features in the remaining two first
families. First, the occupation fraction can take any value from $0.1$
to $1.0$. Second, the halos hosting LAEs all have a narrow mass range
$\Delta M\leq 1.0$, half of them $\Delta M\leq 0.5$.

The existence of models with high occupation fractions $f_{\rm
  occ}>0.2$ is unexpected from previous clustering analysis that do
not take  account the effect of cosmic variance in the same way we
propose in this \documentname. Nevertheless, values of an occupation
fraction $f_{\rm occ}=0.1$ and the mass range predicted by all of our
models are consistent with already published observational results
\citep{Gawiser07,Ouchi2010}. This suggests that explicitly taking
cosmic variance of the observed fields into account (the only point
not included in the analysis already cited) hinders the possibility of
constraining the LAEs occupation fraction.

@@

We proceeded with a thorough exploration of the space of 
parameters ($M_{\rm min}$, $M_{\rm max}$, $f_{\rm occ}$) describing the
model. We look for consistency with two observational constraints: the
surface number density distribution and the angular correlation
function. Out of the initial $9000$  combinations of parameters in the
model we find $40$ combination of parameters consistent with
observations.

@@

The wide range of occupation fraction values is unexpected from
previous studies \citep[i.e.][]{Gawiser2007,Ouchi2010}  that
unanimously favor low values on the range of $10^{-2}$ and only
consider values of $0.1$ as an upper limit. Nevertheless, all the halo
mass ranges deduced in previous analysis are a subset of the models we
find in this paper. All our the best models support the notion that the
most massive halos at $z=3.1$ do not host detectable LAEs.

>>>>

\begin{abstract}
%
  We study the impact of cosmic variance in constraining the mass and
  occupation fraction of dark matter halos hosting \ly Emitting
  Galaxies at high redshift. We use an N-body simulation to construct
  mock fields with the same typical size of observed fields at
  $z=3.1$ to match the observed number density distribution across
  fields and the angular correlation function. In our model a dark
  matter halo with mass in the range $M_{\rm min}<M_{\rm h}<M_{\rm
    max}$ can only host one detectable LAE with a probability $0.1
  \leq f_{\rm occ}\leq 1.0$.  Our analysis shows that the clustering
  and number density information are insufficient to impose a tight
  constraint on the occupation fraction. On the other hand, the
  minimum mass and maximum mass can be constrained to the range
  $M_{\rm  max}<10^{12}$\hMsun\ and $10^{10.2}\hMsun\leq M_{\rm
    min}\leq 10^{11.5}\hMsun$. Additionally, we find that the
  consistent models have a narrow mass range, $\Delta M \equiv \log_{10}M_{\rm max} -
  \log_{10} M_{\rm min}$, smaller than $1.0$ dex, with a majority in
  the range $\Delta M<0.5$ dex. These results fall into constraints
  already presented in the literature. However, the wide range of
  values for the occupation fraction and the narrow mass range $\Delta
  M$ are novel results allowed by taking into account the influence of
  cosmic variance on the statistics derived from observations.  
\end{abstract}

@@

Capitalizing LAEs observations requires an understanding of
their place in the context of a given structure formation model in a
cosmological context. Under the current paradigm the dominant matter
content of the Universe is to be found in dark matter (DM) and each
galaxy is thought to be hosted by a larger dark matter structure known
as a halo. \citep{Peebles1980,SpringelNature05}.

@@

Some theoretical attempts to solve this problem using an  ab-initio
approach. They start from the DM halo population to infer the
intrinsic star formation rates and \ly  luminosities. From these
values they estimate the amount of \ly photons that
escape each galaxy and compute the observed luminosity for each
galaxy. These models can predict different observables including: the
luminosity function, correlation function and the equivalent width
distributions. Such modelling has been implemented in semi-analytic
models \citep{Garel2012,Orsi2012} and  full N-body
hydrodynamical simulations \citep{Laursen2007, Dayal2009,
  ForeroRomero2011, Yajima2012}.

@@

In this paper we investigate the impact of cosmic variance in
constraining the mass and occupation fraction of halos hosting LAEs.
We do this by constructing mock catalogs from cosmological simulations.
Our method populates each halo in the simulation with a LAE without
predicting a \ly  luminosity. This bypasses all the physical
uncertainties associated to star formation and radiative transfer.
We build mock surveys following the geometry of the observed fields by
\cite{Yamada2012} to compare them  against observations in terms of
the number density distribution and the angular correlation
function. This allows us to find a range of parameters of our model
that are consistent with observations while taking into account
cosmic variance.

@@

In the next subsections we describe in detail the four key
elements in our work-flow. First, we present the 
observations we take as a benchmark. Second, we describe the main
characteristics of the N-body simulation and the halo catalogs we
use. Third, we list the important parameters of the simplified
model that we use to populate the halo catalogs with LAEs. Finally, we
describe the statistical tests we adopt to compare observations and mocks.

@@

The narrow band filter used in the survey is
centered at $4977$ \AA~with  $77$ \AA~width, corresponding to the
redshift range $z=3.062$-$3.125$ and $41$ \hMpc comoving scale for the
detection of the Lyman-$\alpha$ line centered at $z=3.09$. The authors
reported a total of $2161$  LAEs with an observed equivalent width, in
the observer frame, larger than $190$ \AA~over a total survey area of
$2.42$ deg$^{2}$ that includes 12 sub-fields,  this corresponds to an
average surface number density of $0.24\pm 0.01$ arcmin$^{-2}$.  The
distribution of the surface number densities is shown in Figure
\ref{fig:number_density}.

@@

The second benchmark is the angular correlation function
(ACF). Unfortunately, \cite{Yamada2012} do  not report an ACF measurement for
their fields. Instead, we use the results by   \cite{Ouchi2008} who
reported the ACF  over a region of $1$ deg$^2$ over the SXDS
field.

@@

We construct a total $5\times 7 \times 6=210$ sub-volumes from a snapshot in the Bolshoi
simulation. In each mock field a LAE is assigned to the position of a
dark matter halo if the halo mass is in the range allowed by the model
$M_{\rm min}<M_{\rm h}<M_{\rm max}$ and a random variable taken from a
homogeneous distribution $0\leq \xi<1$ is smaller than the occupation
fraction $\xi<f_{\rm occ}$.

@@

The lower limit for the parameter $M_{\rm min}$ is set by the minimum
occupation fraction we decide to consider. At $M_{\rm
  min}=10^{10}$\hMsun the halo number density around that mass range
is $\sim 10$ times higher than the observational constraints
for LAEs. This means that models in that mass range and an occupation
fraction $f_{\rm  occ}=0.1$ have the possibility to be compatible with
observations.  Similarly, exploring occupation fractions on the order
of $f_{\rm occ}=0.01$ only makes sense for halo populations in the
mass range of $10^{9}$\hMsun which are abundant enough to reproduce
the observational constraint on the number density. In our case, this
 mass range is unresolved by the Bolshoi simulation and therefore
is not considered in the present study.

@@

The observed and mock ACF are fit to a power-law function:
\begin{equation}
\omega(\theta) = \left(\frac{\theta}{\theta_{0}}\right)^{-\beta}, 
\label{eq:fitting}
\end{equation}
%
where $\theta_0$ and $\beta$ are free parameters. The fit is done
using a least square minimization procedure. For each mock field we
obtain a covariance matrix that gives us the uncertainty in the
parameters $\beta$ and $\theta_0$.  For a model we derive a mean
value and a dispersion computed from the fit of all mock fields. We
consider that a model is consistent with observations if the two
parameters $\beta$ and $\theta_0$ are equal within a $1$-$\sigma$
range.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.46\linewidth,angle=0]{Fig7_corr_params.pdf}  
\includegraphics[width=0.46\linewidth,angle=0]{Fig7_mmin_vs_theta.pdf} 
\end{center}
\caption{Left panel. Values for the free parameters $\theta_{0}$ and $\beta$
in the fitting formula (Eq. \ref{eq:fitting}) for the angular
correlation function. Blue dots correspond to simulations and the
green line to observations by \citet{Ouchi2008,Ouchi2010}. The error
bars in the theoretical data correspond to the quadratic average of
the fitting uncertainties for each mock survey. Right panel. Values of
$\theta_{0}$  and $\log_{10}(M_{\rm min}/M_{\odot} h^{-1})$. Blue and
green point represents the models where  $\log_{10}(M_{\rm max}/M_{\odot}
h^{-1})<12.0$ and  $\log_{10}(M_{\rm min}/M_{\odot} h^{-1})>12.0$
respectively. .  The horizontal gray band represents the observational
constraints on $\theta_{0}$ established by
\citet{Ouchi2008,Ouchi2010}. Green points has been displaced by
$0.01$ dex of its original value to avoid overlapping between blue and
green points.} \label{fig:correlation_parameters}   
\end{figure*}

@@

Figure \ref{fig:halos} allows us to understand why only a specific range of
models ${\mathcal M}$ can be expected to be consistent with
observations. From Figure \ref{fig:halos} we can read that models with
a minimum mass $M_{\rm min}>10^{12}$\hMsun always have a surface
number density lower than the observational constrain, making them
incompatible with observations; there are simply too few halos
compared to observed LAEs. The opposite is true in models with $M_{\rm
  min}<10^{11.0}$\hMsun, which have a surface number density larger
than observations. In those cases the maximum mass $M_{\rm max}$ and the
occupation fraction $f_{\rm occ}<1.0$  can be tuned in order to lower
the halo number density to match observations.

@@

The models in the third family are barely consistent with the
constraints from the ACF. These models have another particular
feature: their minimum mass is exactly $M_{\rm min}=10^{10.9}\hMsun$
and $M_{\rm max}>10^{12.0}\hMsun$ . They have values of $\beta$
close to $1.0$ (instead of the observational value of $0.8$) and
$\theta_0$ on the range of $13^{\prime}$ (the mean observational
value is close to $8^{\prime}$) but are considered consistent
because their large error bars barely overlap with the observational
expectation (see figure \ref{fig:correlation_parameters}).
Because these models are barely compatible with observations and
are a minority of the consistent models, we exclude them from the
discussion.

\begin{figure*}
\begin{center}
\includegraphics[width=0.35\linewidth,angle=0]{Fig6_mass.pdf}
\hspace{5mm}
\includegraphics[width=0.35\linewidth,angle=0]{Fig6_f_occ.pdf}
\end{center} 
\caption{$M_{\rm min}$-$M_{\rm max}$ (left) and $M_{\rm
    min}$-$f_{\rm occ}$ (right) planes showing the models fulfilling both
   constraints on the maximal number of consistent mocks and the
  angular correlation function from \citet{Ouchi2008,Ouchi2010}.   
  \label{fig:restriction_mock_and_f_occ_corr}} 
\end{figure*}

@@

There are two interesting features in the remaining two first
families. First, the occupation fraction can take any value from $0.1$
to $1.0$. Second, all constrained models of halos hosting LAEs 
 have a narrow mass range $\Delta M\leq 1.0$, and half of them 
 have $\Delta M\leq 0.5$.

The existence of models with high occupation fractions $f_{\rm
  occ}>0.2$ is unexpected from previous clustering analysis that do
not take  into account the effect of cosmic variance in the same way we
propose in this \documentname. Nevertheless, values of an occupation
fraction $f_{\rm occ}=0.1$ and the mass range predicted by all of our
models are consistent with already published observational results
\citep{Gawiser07,Ouchi2010}. This suggests that explicitly taking
cosmic variance of the observed fields into account (the only point
not included in the analysis already cited) hinders the possibility of
constraining the LAEs occupation fraction.

@@

We proceeded with a thorough exploration of the space of 
parameters ($M_{\rm min}$, $M_{\rm max}$, $f_{\rm occ}$) describing the
model. We look for consistency with two observational constraints: the
surface number density distribution and the angular correlation
function. Out of the initial $9000$  combinations of parameters in the
model we find $40$ combination of parameters consistent with
observations.  The minimum mass and maximum mass of these consistent
 models are found to be in the ranges:   $M_{\rm  max}<10^{12}$\hMsun\ 
 and $10^{10.2}\hMsun\leq M_{\rm min}\leq 10^{11.5}\hMsun$.

@@

The wide range of occupation fraction values is unexpected from
previous studies \citep[i.e.][]{Gawiser2007,Ouchi2010}  that
unanimously favor low values on the range of $10^{-2}$ and only
consider values of $0.1$ as an upper limit. Nevertheless, all the halo
mass ranges deduced in previous analysis are a subset of the models we
find in this paper. All our  best models support the notion that the
most massive halos at $z=3.1$ do not host detectable LAEs.
